A Monte Carlo Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Dump of nse07 paper into tex file

>>>>

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Matrix Solution Methods}
\label{sec:monte-carlo-matrix}

Consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\
    &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
spectral radius, $\rho(\vH)$, of the iteration matrix satisfies
\begin{equation}
  \rho(\vH) < 1\:,
\end{equation}
we can expand $\vA^{-1}$ using the \latin{Neumann Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} =  \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\
    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
\forall\ \vb \in \mathcal{R}^{N}$~\cite{kelley_1995}.

\subsection{Direct Method}
\label{sec:direct-method}

Now we consider a Monte Carlo method that can be used to estimate the solution
to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First, rewrite
Eq.~(\ref{eq:neumann_series}) in order to calculate a component of
$\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
This series can be interpreted as a series of transitions from
$i_{m-1}\rightarrow i_m$ that can be simulated by a random walk.
Let $X$ be a random variable sampled from a random walk with $k$
events that initiates in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
and each particle starts with unit weight. When the particle
transitions between states, for example $i\rightarrow j$, the weight
is multiplied by the factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:weight}
\end{equation}
Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
    &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.
Therefore, the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased
estimator of the components of $\vx$ provided $\rho(\vH) < 1$.

We are left to define the transition probabilities, $p_{ij}$. The most
straightforward approach is to set
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
  \label{eq:probability}
\end{equation}
Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function for selecting a new state $j$,
given that the current state is $i$.  Random walks can be terminated
in two ways: the matrix can be augmented with a terminating event
equation that describes the probability that a particle ends its walk,
or the random walk can be terminated by weight cutoff.  Generally, we
choose to terminate random walks using a weight cutoff as opposed to
augmenting the matrix.

\subsection{Adjoint Method}
\label{sec:adjoint-method}

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk.  In
the mathematical literature this is referred to as an adjoint method;
however, to transport practitioners this approach more closely
resembles a traditional forward method.  In this method, the weight
change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are
normalized over a column, as opposed to the forward method in which
the probabilities are normalized over a row.  This is equivalent to
forming the Neumann series in reverse order.  Correspondingly, the
estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
                                  w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source in cell/state $i_0$.  The
Kronecker delta implies that tallies are only made in the cell/state
where the random walk currently resides.  This is the common approach
in standard Monte Carlo transport simulations.  We will investigate
this approach as applied to the 1T thermal radiation diffusion
problem in \S~\ref{sec:solut-radi-diff}.

Similar to the Direct Method, the Adjoint Method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that
\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}
where $W_f$ is the terminating weight of the random walk, $W_c$ is the
input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
weight of the source particle.  The random walk ends on the
$m^\text{th}$ step if $W_m < W_f$.

%%---------------------------------------------------------------------------%%
\section{Synthetic-Acceleration Monte Carlo}
\label{sec:iter-refin-monte}

The methods presented in \S~\ref{sec:monte-carlo-matrix} are
characterized by slow convergence rates.
Halton~\cite{halton_1962,halton_1994} proposed a staged residual
scheme called Sequential Monte Carlo to speed up the convergence of
these methods.  A variant of this scheme has been successfully applied
to the 1D nonlinear thermal radiation diffusion equation in
Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:
\begin{subequations}
  \begin{gather}
    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the adjoint method is used to estimate the solution to
Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
convergence.  This iteration sequence is closely related to iterative
refinement; the exception being that there is no update to to
$x^{l+1}$ between residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration scheme for fixed-point iteration. The
Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method is
defined as follows:
\begin{subequations}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:, \\
    %%
    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,\\
    %%
    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
    %%
    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:general-MCSA}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  here.}.  The residual is calculated and is used as a source to
estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method.
The error is used to calculate an updated iterate of the solution
vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence
based on the following stopping criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

%%---------------------------------------------------------------------------%%
\section{Discrete Form of the Radiation Diffusion Equation}
\label{sec:discr-form-radi}

\subsection{Equilibrium Diffusion Model}

The general form of the equilibrium diffusion equation is
\cite{morel_1996}
\begin{equation} 
  (\rho\Cv + 4aT^{3})\pder{T}{t} -
  \nabla\cdot{\Bigl(\frac{4acT^{3}}{3\ros}\Bigr)\grad{T}} = Q\:,
  \label{eq:eq-diff-T}
\end{equation}
where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
$[\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
heat capacity of the material, $\rho\equiv\rho(\ve{r})$
$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
$a=0.01372$ $[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
the radiation constant, $c=299.79$ $[\text{cm}\cdot\text{sh}^{-1}]$ is
the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
the temperature that characterizes both the radiation and the
material.  The opacity, $\ros\equiv\ros$(\ve{r},T(\ve{r},t))
$[\text{cm}^{-1}]$, is the Rosseland Mean opacity.  The source,
$Q\equiv Q(\ve{r},t)$
$[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{sh}^{-1}]$, is assumed to
be independent of $T$.

We chose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{Jerks}\cdot\text{cm}^{-2}\cdot\text{sh}^{-1}]$, as the primary
state variable.  The intensity is defined,
\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t) \\
    &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\
    &= acT^{4}\:,
  \end{split}
\end{equation}
where $\psi$ is the angular radiation intensity.  The Planck function
(or Planckian) is defined as,
\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}
where $h$ is Planck's constant and $k$ is the Boltzmann constant.
Utilizing this definition of $\phi$ we can write
\begin{equation}
  \label{eq:partial-T}
  \partial T = \frac{1}{4acT^{3}}\partial\phi\:.
\end{equation}
Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} - 
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}
where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined
\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}
Equation~\eqref{eq:eq-diff} is the model equation we will solve using
the Monte Carlo techniques presented in
\S~\ref{sec:monte-carlo-matrix} and \ref{sec:iter-refin-monte}.

\subsection{Discrete Diffusion Equation}

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff})
and lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and 
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\
  \qn &= \sign\phin + Q^n\:,\\
  \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance
equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Fig.~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.  We now apply the divergence
theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\
    &= \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk \\
    &+ (F_{k+1/2} - F_{k-1/2})\Di\Dj
    + \sign_{ijk}\phi_{ijk}V_{ijk} = \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

Using Fick's Law, Eq.~(\ref{eq:fick}), we can write $O(\Delta^2)$
expressions for the face-centered fluxes.  Defining $l\in\{i,j,k\}$,
the face-centered fluxes are
\begin{align}
  F_{l+1/2} &= -D_{l+1/2}\frac{\phi_{l+1} -
    \phi_{l}}{\Delta_{l+1/2}}\:, \label{eq:F_l+1/2}\\
  F_{l-1/2} &= -D_{l-1/2}\frac{\phi_{l} -
    \phi_{l-1}}{\Delta_{l-1/2}}\:, \label{eq:F_l-1/2}
\end{align}
where
\begin{align}
  \Delta_{l+1/2} &= \frac{\Delta_{l+1} + \Delta_{l}}{2}\:,\\
  \Delta_{l-1/2} &= \frac{\Delta_{l} + \Delta_{l-1}}{2}\:.
\end{align}
In order to be conservative, the flux must be continuous across a
face.  Thus, the face-centered flux evaluated from either side of the
face must be equivalent,
\begin{equation}
  \begin{aligned}
    F_{l+1/2}^{+} &= F_{l+1/2}^{-}\:,\\
    2D_{l+1}\frac{\phi_{l+1} - \phi_{l+1/2}}{\Delta_{l+1}} &=
    2D_{l}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,\\
    \frac{1}{3\sigma_{l+1}}\frac{\phi_{l+1} - \phi_{l+1/2}}{\Delta_{l+1}} &=
    \frac{1}{3\sigma_{l}}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,
  \end{aligned}
\end{equation}
where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
clarity.  Solving this system for $\phi_{l+1/2}$ and plugging into
either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
expression equal to Eq.~(\ref{eq:F_l+1/2}), one finds the
face-centered diffusion coefficient,
\begin{equation}
  D_{l+1/2} = \frac{2\Delta_{l+1/2}}{3\sigma_l\Delta_l + 
    3\sigma_{l+1}\Delta_{l+1}}\:.
  \label{eq:D_l+1/2}
\end{equation}
Following the same procedure at the $l-1/2$ face gives the diffusion
coefficient on the low-side face,
\begin{equation}
  D_{l-1/2} = \frac{2\Delta_{l-1/2}}{3\sigma_l\Delta_l + 
    3\sigma_{l-1}\Delta_{l-1}}\:.
  \label{eq:D_l-1/2} 
\end{equation}
Using Eqs.~(\ref{eq:F_l+1/2}), (\ref{eq:F_l-1/2}), (\ref{eq:D_l+1/2}),
and (\ref{eq:D_l-1/2}) in the discrete flux-balance equation,
(\ref{eq:flux-balance-difference}), we obtain
\begin{equation}
  \begin{aligned}
    \sigT\phi_{ijk} 
    - \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} &
    - \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}
    - \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
    &- \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k} 
    - \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
    - \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
    = \qn_{ijk}\:.
  \end{aligned}
  \label{eq:discrete-diffusion}
\end{equation}
Here,
\begin{equation}
  \sigT = \sigp_i + \sigm_i + \sigp_j + \sigm_j + \sigp_k + \sigm_k
  + \sign_{ijk}\:,
\end{equation}
and
\begin{subequations}
  \begin{align}
    \sigp_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di
      + 3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\
    \sigm_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di
      + 3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\
    \sigp_{j} &= \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj
      + 3\sigma_{i\,j+1\,k}\Delta_{j+1}}\:,\\
    \sigm_{j} &= \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj
      + 3\sigma_{i\,j-1\,k}\Delta_{j-1}}\:,\\
    \sigp_{k} &= \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk
      + 3\sigma_{ij\,k+1}\Delta_{k+1}}\:,\\
    \sigm_{k} &= \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk
      + 3\sigma_{ij\,k-1}\Delta_{k-1}}\:. 
  \end{align}
  \label{eq:sigmas}
\end{subequations}
Also, note that
\begin{align*}
  \sigp_{i-1} &= \frac{1}{\Delta_{i-1}}\frac{2}{3\sigma_{ijk}\Di
    + 3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\
  \sigm_{i+1} &= \frac{1}{\Delta_{i+1}}\frac{2}{3\sigma_{ijk}\Di
    + 3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\
  \ldots\:.
\end{align*}
Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas})
represent the discretized, 1T diffusion equation on an orthogonal
3D grid.

Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas}) have
been written in a manner that is amenable to Monte Carlo
interpretation.  One can envision the terms in Eq.~(\ref{eq:sigmas})
representing leakage out of each face of a cell.  The off-diagonal
terms then represent sources into the cell, or, equivalently, these
terms are leakage out of the neighboring cells.  A limitation of this
interpretation is that it requires a symmetric system to implement the
Monte Carlo scheme.  More specifically, the operator $\ve{D}$ must be
an H-matrix \cite{kelley_1995} such that the off-diagonal elements are
negative and symmetric.  In other words, the leakage out of a cell
must be equivalent to the source into the adjacent cell.

The methods described in \S\S~\ref{sec:monte-carlo-matrix} and
\ref{sec:iter-refin-monte} do not suffer this limitation.  Thus, we
will interpret the solution via MCSA as a transport-like process, but
we do not require any constraints on the system other than the
spectral radius must be less than 1.

Before completing the derivation of discrete equations required to
solve Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary
conditions.  For this work, the Marshak Boundary condition suffices,
\begin{equation}
  f_b^\pm = \frac{1}{4}\phi - \frac{1}{2}\ve{F}\cdot\hat{\ve{n}}\:,\quad
  \vr\in \partial\Gamma\:,
\end{equation}
where $f_b$ is the incoming partial current on the problem boundary,
$\partial\Gamma$.  On any low boundary we can write the flux using a
first-order discrete form of Fick's Law,
\begin{equation}
  F_{1/2} = -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1}\:.
  \label{eq:low-boundary-flux}
\end{equation}
As before, we enforce continuity of flux on the boundary using the
Marshak Boundary condition,
\begin{equation}
  -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1} = 2f_b^{+}
  -\frac{1}{2}\phi_{1/2}\:. 
\end{equation}
Solving for $\phi_{1/2}$ and plugging into
Eq.~(\ref{eq:low-boundary-flux}) yields an expression for the low-side
boundary flux,
\begin{equation}
  F_{1/2} = \frac{2}{3\sigma_1\Delta_1 + 4}(4f_b^{+}-\phi_1)\:.
\end{equation}
Following an identical procedure on high-side problem boundaries gives
\begin{equation}
  F_{N_l+1/2} = \frac{2}{3\sigma_{N_l}\Delta_{N_l} +
    4}(\phi_{N_l}-4f_b^{-})\:, 
\end{equation}
where $l\in\{i,j,k\}$ is defined over the range $[1,N_l]$.

These boundary fluxes take the place of the interior face-centered
fluxes (defined in Eqs.~(\ref{eq:F_l+1/2}) and (\ref{eq:F_l-1/2})) in
Eq.~(\ref{eq:flux-balance-difference}) yielding the following
modifications to Eqs.~(\ref{eq:sigmas}) on a boundary,
\begin{align}
  \text{low boundary:}\quad & \sigma_1^{-} = \frac{1}{\Delta_1}
  \frac{2}{3\sigma_1\Delta_1 + 4}\:,\\
  \text{high boundary:}\quad & \sigma_{N_l}^{+} = \frac{1}{\Delta_{N_l}}
  \frac{2}{3\sigma_{N_l}\Delta_{N_l} + 4}\:.
\end{align}
Furthermore, the following contributions are added to the source,
$\qn_{ijk}$ on a boundary cell,
\begin{align}
  \text{low boundary:}\quad & \frac{8}{\Delta_1}
  \frac{1}{3\sigma_1\Delta_1 + 4}f_b^{+}\:,\\
  \text{high boundary:}\quad & \frac{8}{\Delta_{N_l}}
  \frac{1}{3\sigma_{N_l}\Delta_{N_l} + 4}f_b^{-}\:.
\end{align}
Note that for vacuum boundary conditions, $f_b^{\pm} = 0$.

For reflective boundary conditions we have the simple modification
that
\begin{align}
  F_{1/2} &= 0\:, \\
  F_{N_l+1/2} &= 0\:.
\end{align}
Thus, there are no leakage or adjacent cell source terms resulting
from this boundary face.

\subsection{Preconditioning}
\label{sec:preconditioning}

The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can
be written in matrix-operator form as follows,
\begin{equation}
  \ve{D}\bphi = \ve{q}\:,
  \label{eq:diffusion-matrix}
\end{equation}
where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we
must precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the
MCSA method.  Preconditioning accomplishes two purposes: it reduces
the spectral radius of $\ve{D}$ so that Monte Carlo solution methods
are applicable, and it reduces the number of iterations required for
convergence.  We apply left-preconditioning and write
\begin{equation}
  \ve{M}^{-1}\ve{D}\bphi = \ve{M}^{-1}\ve{q}\:.
  \label{eq:preconditioned-diffusion}
\end{equation}
A good choice for the preconditioner, $\ve{M}$, is
\begin{equation}
  \ve{M}=\diag(\ve{D})\:.
  \label{eq:preconditioner}
\end{equation}
Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$
accomplishes several objectives:
\begin{itemize}
\item It reduces the spectral radius such that
  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
  fewer iterations are required resulting in accelerated convergence.
\item Makes the diagonal elements of the iteration matrix zero,
  $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no time
  wasted in transitions $i\rightarrow i$ during the random walk
  process.
\end{itemize}
The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of
any additional preconditioner choices.

A side effect of this preconditioner choice is that the preconditioned
system, $\ve{M}^{-1}\ve{D}$, is no longer SPD.  This situation can be
rectified by doing right-left preconditioning and defining
\begin{equation}
  \ve{M} = \ve{L}\ve{L}^{T}\:,
\end{equation}
where $\ve{L} = \ve{L}^{T} = \sqrt{\ve{M}}$.  Applying right-left
preconditioning gives
\begin{equation}
  \begin{gathered}
    \ve{M}^{-1}\ve{D}\ve{M}^{-1}\ve{M}\bphi = \ve{M}^{-1}\ve{q}\:,\\
    (\ve{LL}^T)^{-1}\ve{D}(\ve{LL}^T)^{-1}\ve{LL}^T\bphi =
    (\ve{LL}^T)^{-1}\ve{q}\:, \\
    \ve{L}^{-T}\ve{L}^{-1}\ve{D}\ve{L}^{-T}\ve{L}^T\phi =
    \ve{L}^{-T}\ve{L}^{-1}\ve{q}\:,\\ 
    (\ve{L}^{-1}\ve{D}\ve{L}^{-T})\ve{L}^T\bphi = \ve{L}^{-1}\ve{q}\:,\\
    \hat{\ve{D}}\hat{\bphi} = \hat{\ve{q}}\:.
  \end{gathered}
  \label{eq:right-left-preconditioning}
\end{equation}
This type of preconditioning is also referred to as row-column
scaling.  The resulting preconditioned matrix, $\hat{\ve{D}}$, is now
SPD, and we can use Conjugate Gradient (CG) solvers.  Therefore, we
will use this preconditioning scheme for deterministic solutions of
Eq.~(\ref{eq:diffusion-matrix}).  However, given that the convergence
properties of $\hat{\ve{D}}$ are similar to $\ve{M}^{-1}\ve{D}$ and
the Monte Carlo solvers do not require SPD matrices, we will be
content with solving Eq.~(\ref{eq:preconditioned-diffusion}) using
MCSA.

The preconditioned system we will solve is defined in
Eq.~(\ref{eq:preconditioned-diffusion}).  Applying $\ve{M}^{-1}$ to
Eq.~(\ref{eq:discrete-diffusion}) yields
\begin{equation}
  \begin{aligned}
    \phi_{ijk} 
    - \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}
    - \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}
    - \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
    &- \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k} 
    - \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
    - \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
    = \frac{\qn_{ijk}}{\sigT}\:.
  \end{aligned}
  \label{eq:preconditioned-discrete-diffusion}
\end{equation}
In the next section we will derive an MCSA method to solve this
equation.

%%---------------------------------------------------------------------------%%
\section{Solution of the Radiation Diffusion Equation}
\label{sec:solut-radi-diff}

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solution steps:
\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}
For clarity, we rewrite the iteration scheme in
Eqs.~(\ref{eq:general-MCSA}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),
\begin{align*}
  %%
  \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
  %%
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\
  %%
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
  %%
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
describe each of these steps in turn.

First, we use Eq.~(\ref{eq:preconditioned-discrete-diffusion}) to
evaluate $\bphi^{l+1/2}$ via a single fixed-point iteration as
follows:
\begin{equation}
  \begin{aligned}
    \phi_{ijk}^{l+1/2} =  
    \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}^l
    + \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l
    +
    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l \\ 
    &+ \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l 
    + \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
    + \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
    + \frac{\qn_{ijk}}{\sigT}\:.
  \end{aligned}
\end{equation}
This result is used to compute the residual of the system.  The
residual then becomes the source for the Monte Carlo simulation.

The adjoint Monte Carlo method described in
\S~\ref{sec:adjoint-method} is used to estimate
\begin{equation}
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
\end{equation}
As noted above, the residual acts as the source for this simulation,
and the operator $\ve{M}^{-1}\ve{D}$ is implicitly defined in
Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte
Carlo interpretations that can be applied to this system.  The first
follows the mathematical presentation given in
\S~\ref{sec:monte-carlo-matrix}.  Namely, we form probabilities from
the iteration matrix and perform random walks to generate unbiased
estimates of the solution vector, $\delta\bphi^{l+1/2}$, using the
estimator in Eq.~(\ref{eq:adjoint-tally}).
